---
title: "SSWEEP_proj_3_18_26"
format: html
---

```{r}
library(tidyverse)
library(janitor)
library(sf)
library(terra)
library(ranger)
library(ggplot2)
library(scales)
library(leaflet)
library(readxl)
library(tidyverse)
library(sf)
library(ggplot2)
library(readxl)
library(terra)
library(ranger)
library(viridis)
library(janitor)
library(randomForest)
library(purrr)
library(reshape2)
library(leaflet)
library(scales)
library(htmltools)
library(vegan)

```
###### Load Data

```{r load data, message=FALSE, warning=FALSE}
##### Load data
  #load data:
  #new weed ID data (xlsx)
  weed_id_tax <- read_excel("../data/SSWEEP_data_with_tax_and_indices_02_2026.xlsx", sheet = "SSWEEP")
  #indices:
  weed_idx <- read_excel("../data/SSWEEP_data_with_tax_and_indices_02_2026.xlsx", sheet = "Calculations for each grid")

  #Weed ID data (xlsx):
  weed_id <- read_csv("../data/SSWEEP_weed_id_count_data.csv")
    #keep Total sheet but discard other sheets:
    
  #surface sampling (15cm) (.csv):
  surface_sampling <- read_csv("../data/SSWEEP_surface_SP_samples.csv")
  
  #full profile samples:
  # full_profile <- read_csv("../data/SSWEEP_full_profile_samples.csv")
    #covariate stack (.tfs):
     # cov_stack <- rast("../data/covariate_stack_3m.tif")

```

##### Data wrangling: join weed ID data to surface sampling points

```{r data wrangling, message=FALSE, warning=FALSE}
#clean weed_id_tax:
weed_id_tax <- weed_id_tax %>%
  select(1:143) %>%
  #select rows 1:72:
  slice(1:72)

#weed id columns are rows under surface sampling row_num column:
#pivot all columns of weed id:
weed_id_long <- weed_id_tax %>%
  pivot_longer( cols = -c("Common Name","Genus","Species","Family","Life Cycle","Season","Type"),
               names_to = "row_num", 
               values_to = "weed_count") %>%
  #rename ...1 to weed_species 
  rename(weed_species = "Common Name")

#join row_num 
surface_sampling_weed <- surface_sampling %>%
  left_join(weed_id_long, by = "row_num") %>%
  clean_names()

head(surface_sampling_weed)

```

### Map of dominant weed family + dominance share (family abundance / total abundance) at each surface sampling point

###### With circle size scaled by total abundance and fill opacity scaled by dominance share. Use an imagery basemap and a color palette for families. Include legends for both size and opacity.

```{r weed map, warning=FALSE}


# ---- 1) dominant family + dominance share ----
weed_point_family <- surface_sampling_weed %>%
  mutate(
    weed_count = as.numeric(weed_count),
    family = str_squish(family)
  ) %>%
  filter(!is.na(weed_count)) %>%
  group_by(row_num) %>%
  mutate(total_abundance = sum(weed_count, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(row_num, family) %>%
  summarise(
    x = first(x),
    y = first(y),
    total_abundance = first(total_abundance),
    family_abundance = sum(weed_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(row_num) %>%
  slice_max(order_by = family_abundance, n = 1, with_ties = FALSE) %>%
  mutate(dominance = family_abundance / pmax(total_abundance, 1)) %>%
  ungroup()

weed_sf <- st_as_sf(weed_point_family, coords = c("x","y"), crs = 4326)

# ---- 2) palettes + scaling ----
fam_levels <- sort(unique(weed_sf$family))
pal <- colorFactor(palette = "Dark2", domain = fam_levels)   # or "Dark2", "Paired", etc.

# radius from total abundance (cap so one crazy point doesn't take over)
rad <- rescale(weed_sf$total_abundance,
               to = c(4, 18),
               from = c(min(weed_sf$total_abundance, na.rm=TRUE),
                        quantile(weed_sf$total_abundance, 0.95, na.rm=TRUE)))
rad <- pmax(4, rad)

# alpha from dominance (keep visible even if low)
alpha_fill <- pmax(0.25, pmin(1, weed_sf$dominance))

# popup text
popup_txt <- paste0(
  "<b>Grid:</b> ", weed_sf$row_num,
  "<br><b>Dominant family:</b> ", weed_sf$family,
  "<br><b>Total abundance:</b> ", weed_sf$total_abundance,
  "<br><b>Family abundance:</b> ", weed_sf$family_abundance,
  "<br><b>Dominance share:</b> ", round(weed_sf$dominance, 2)
)

# ---- 3) leaflet map with imagery basemap ----
# build an "alpha legend" as HTML
alpha_vals <- round(seq(max(0.25, min(weed_sf$dominance, na.rm=TRUE)),
                        max(weed_sf$dominance, na.rm=TRUE),
                        length.out = 4), 2)

alpha_legend <- tags$div(
  style = "
    background: rgba(255,255,255,0.9);
    padding: 8px 10px;
    border-radius: 6px;
    box-shadow: 0 1px 4px rgba(0,0,0,0.3);
    font-size: 12px;
    line-height: 16px;
  ",
  tags$div(tags$b("Dominance (fill opacity)")),
  tags$div(style="margin-top:6px;",
    lapply(alpha_vals, function(a){
      tags$div(style="display:flex; align-items:center; margin-bottom:4px;",
        # a little circle swatch with different alpha
        tags$span(style = sprintf(
          "display:inline-block; width:12px; height:12px; border-radius:50%%;
           background: rgba(0,0,0,%.2f); border:1px solid #555; margin-right:8px;",
          a
        )),
        tags$span(sprintf("%.2f", a))
      )
    })
  ),
  tags$div(style="margin-top:6px; color:#444;",
           "More opaque = stronger dominance")
)
# ---- SIZE LEGEND (abundance -> circle radius) ----
# pick 4 representative abundance values (min, median, 75th, 95th-ish)
ab_vals <- c(
  min(weed_sf$total_abundance, na.rm = TRUE),
  median(weed_sf$total_abundance, na.rm = TRUE),
  quantile(weed_sf$total_abundance, 0.75, na.rm = TRUE),
  quantile(weed_sf$total_abundance, 0.95, na.rm = TRUE)
) %>% as.numeric() %>% unique() %>% sort()

# convert those abundance values to radii using the SAME mapping you used for points
rad_fun <- function(x) {
  rescale(x, to = c(4, 18),
          from = c(min(weed_sf$total_abundance, na.rm=TRUE),
                   quantile(weed_sf$total_abundance, 0.95, na.rm=TRUE))) %>%
    pmax(4) %>% pmin(18)
}

rad_vals <- rad_fun(ab_vals)

size_legend <- tags$div(
  style = "
    background: rgba(255,255,255,0.9);
    padding: 8px 10px;
    border-radius: 6px;
    box-shadow: 0 1px 4px rgba(0,0,0,0.3);
    font-size: 12px;
    line-height: 16px;
  ",
  tags$div(tags$b("Total abundance (size)")),
  tags$div(style="margin-top:6px;",
    lapply(seq_along(ab_vals), function(i){
      r <- rad_vals[i]
      tags$div(style="display:flex; align-items:center; margin-bottom:6px;",
        # circle swatch sized by radius (diameter = 2r)
        tags$span(style = sprintf(
          "display:inline-block; width:%dpx; height:%dpx; border-radius:50%%;
           background: rgba(0,0,0,0.25); border:1px solid #555; margin-right:10px;",
          round(2*r), round(2*r)
        )),
        tags$span(format(round(ab_vals[i], 1), trim = TRUE))
      )
    })
  )
)

leaflet(weed_sf) %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "Light") %>%
  addLayersControl(baseGroups = c("Imagery", "Light"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addCircleMarkers(
    radius = rad,
    color = ~pal(family),
    weight = 2,
    opacity = 1,
    fillColor = ~pal(family),
    fillOpacity = alpha_fill,
    popup = popup_txt
  ) %>%
  addLegend("bottomright",
            pal = pal,
            values = ~family,
            title = "Dominant family",
            opacity = 1) %>%
  addControl(alpha_legend, position = "bottomleft") %>%
  addControl(size_legend, position = "topleft")

```

```{r}

# re-read sheet 1 cleanly (so we definitely have the bottom rows)
ssweep <- read_excel("../data/SSWEEP_data_with_tax_and_indices_02_2026.xlsx",
                     sheet = "SSWEEP")

# grid columns are like 1A, 22C, 93H, etc.
grid_cols <- names(ssweep)[str_detect(names(ssweep), "^\\d+[A-Za-z]+$")]

# rows at the bottom of sheet 1 that you want
idx_rows <- c(
  "Simpson's diversity index",
  "Shannon Weiner diversity index",
  "Pielou's evenness index",
  "Berger-Parker dominance index"
)

# extract ONLY those index rows + their values across the grid columns
idx_by_point <- ssweep %>%
  filter(`Common Name` %in% idx_rows) %>%
  mutate(idx_name = make_clean_names(`Common Name`)) %>%
  mutate(idx_name = make.unique(idx_name)) %>%   # handles the duplicate Berger-Parker row
  select(idx_name, all_of(grid_cols)) %>%
  pivot_longer(all_of(grid_cols), names_to = "row_num", values_to = "value") %>%
  mutate(value = as.numeric(value)) %>%
  pivot_wider(names_from = idx_name, values_from = value) %>%
  clean_names()

# join onto your surface points (now you have indices per point)
surface_sampling_idx <- surface_sampling %>%
  mutate(row_num = as.character(row_num)) %>%
  left_join(idx_by_point, by = "row_num") %>%
  clean_names()

head(surface_sampling_idx)

```

```{r}
#plot leaflet of shannon index across surface sampling points:
surface_sampling_idx <- surface_sampling_idx %>%
  select(-berger_parker_dominance_index) %>%
  rename(berger_parker_dominance_index = berger_parker_dominance_index_2)

idx_sf <- st_as_sf(surface_sampling_idx, coords = c("x","y"), crs = 4326)

# ---- 1) find all index columns (adjust pattern if needed) ----
idx_cols <- names(idx_sf) %>%
  keep(~ str_detect(.x, "simpson|shannon|pielou|evenness|berger|parker|dominance"))

idx_cols
# (sanity check) if you want to force a specific order:
idx_cols <- idx_cols[idx_cols %in% c(
  "simpsons_diversity_index",
  "shannon_weiner_diversity_index",
  "pielous_evenness_index",
  "berger_parker_dominance_index"
)]

# ---- 2) popup that ALWAYS includes all indices ----
popup_all <- purrr::pmap_chr(
  idx_sf %>% st_drop_geometry() %>% select(row_num, all_of(idx_cols)),
  function(...) {
    r <- list(...)
    rn <- r$row_num
    r$row_num <- NULL

    lines <- purrr::imap_chr(r, \(v, nm) {
      if (is.null(v) || is.na(v)) return(paste0("<b>", nm, ":</b> NA"))
      paste0("<b>", nm, ":</b> ", round(as.numeric(v), 4))
    })

    paste0("<b>Row:</b> ", rn, "<br>", paste(lines, collapse = "<br>"))
  }
)

# Precompute coordinates once
coords <- st_coordinates(idx_sf)

# ---- 3) build leaflet with one overlay per index ----
m <- leaflet(idx_sf) %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "Light")

for (col in idx_cols) {
  pal <- colorNumeric("viridis", domain = idx_sf[[col]], na.color = "transparent")

  m <- m %>%
    addCircleMarkers(
      lng = coords[,1], lat = coords[,2],
      radius = 7,
      stroke = TRUE, weight = 1, color = "white",
      fillColor = pal(idx_sf[[col]]),
      fillOpacity = 0.9,
      popup = popup_all,
      group = col
    ) %>%
    addLegend(
      position = "bottomright",
      pal = pal,
      values = idx_sf[[col]],
      title = col,
      group = col
    )
}

m %>%
  addLayersControl(
    baseGroups = c("Imagery", "Light"),
    overlayGroups = idx_cols,
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(idx_cols[-1])  # show first index layer by default
```








### Build training table for RF soil-property rasters

###### Based on surface samples, using covariates extracted at soil sampling points.

```{r RF soil-property rasters, message=FALSE, warning=FALSE}
#questions:
  #can any soil properties predict weed species presence/abundance or diversity indices?
  #what topographic covariates are most predictive of weed species?
  #are there particular soil nutrients that are associated with particular weed species or diversity indices or weed taxa?
  #are there particular disturbance covariates that are associated with particular weed species or diversity indices or weed taxa?
  #are there particular full profile summaries (e.g. root depth) that are associated with particular weed species 
    #or diversity indices or weed taxa?


#plan: pull in disturbance covariate rasters and potentially create rasters such as distance to roads
#pull in other covariates
#potentially pull in full profile data (such as root depth data) to use as predictors in RF soil-property rasters (but only if we have enough points with root depth data to make it worthwhile)
#merge with soil nutrients
#correlation matrix between taxa, nutrients
#correlation matrix between disturbance covariates and weed species and indices
#correlation matrix between full profile summaries and weed species/indicies/nutrients



# Disturbance raster (UTM 17N)
current_dist <- rast("../data/disturbance/currently_dist_binary.tif")

# Recode categorical -> binary (you already did this)
lev <- levels(current_dist)[[1]]
id_col <- names(lev)[1]
yes_id <- lev[[id_col]][lev$CurrentDis == "Yes"]
current_dist_bin <- ifel(current_dist == yes_id, 1, 0)
names(current_dist_bin) <- "dist_current"

# OPTIONAL: distance-to-disturbance raster (meters)
# (distance from each cell to nearest disturbed cell)
dist_to_dist <- distance(current_dist_bin, value = 1)
names(dist_to_dist) <- "dist_to_current_m"

# --- disturbance rasters already made ---
# current_dist_bin (0/1) and dist_to_dist (meters)

# points in raster CRS
soil_pts_ll <- surface_sampling %>%
  select(row_num, x, y, lbc, lbc_eq, p_h_2, ca, k, mg, mn, p, zn) %>%
  st_as_sf(coords = c("x","y"), crs = 4326)

soil_pts_utm <- st_transform(soil_pts_ll, crs(current_dist_bin))

#-----------------------------------
# keep projected coordinates
#-----------------------------------
soil_pts_utm_coords <- soil_pts_utm %>%
  mutate(
    x_utm = st_coordinates(geometry)[,1],
    y_utm = st_coordinates(geometry)[,2]
  ) %>%
  st_drop_geometry() %>%
  janitor::clean_names()

# disturbance values already extracted above:
# dist_vals
# dist_dvals

point_base <- soil_pts_utm_coords %>%
  bind_cols(dist_vals, dist_dvals) %>%
  janitor::clean_names() %>%
  mutate(row_num = as.character(row_num))

# point_env now includes projected coords
point_env <- point_base %>%
  left_join(idx_keep, by = "row_num") %>%
  select(
    row_num,
    x_utm,
    y_utm,
    dist_current,
    dist_to_current_m,
    all_of(intersect(soil_vars, names(.))),
    all_of(covar_vars),
    simpsons_diversity_index,
    shannon_weiner_diversity_index,
    pielous_evenness_index,
    berger_parker_dominance_index
  )

analysis_dat <- point_env %>%
  left_join(weed_traits_point, by = "row_num") %>%
  mutate(
    dist_current = as.factor(dist_current)
  )

glimpse(analysis_dat)
# extract disturbance values
dist_vals <- terra::extract(current_dist_bin, vect(soil_pts_utm)) %>% select(-ID)
dist_dvals <- terra::extract(dist_to_dist, vect(soil_pts_utm)) %>% select(-ID)

soil_dist <- soil_pts_utm %>%
  st_drop_geometry() %>%
  bind_cols(dist_vals, dist_dvals) %>%
  clean_names()




# current_dist_bin is 0/1
# make a target raster: 1 where disturbed, NA elsewhere
target <- ifel(current_dist_bin == 1, 1, NA)

# distance to nearest target cell (meters, since UTM)
dist_to_dist <- distance(target)
names(dist_to_dist) <- "dist_to_current_m"

plot(dist_to_dist)
global(dist_to_dist, "range", na.rm=TRUE)
dist_dvals <- terra::extract(dist_to_dist, vect(soil_pts_utm)) %>% select(-ID)

soil_dist <- soil_pts_utm %>%
  st_drop_geometry() %>%
  bind_cols(dist_vals, dist_dvals) %>%
  janitor::clean_names()
```


```{r RF soil-property rasters, message=FALSE, warning=FALSE}

#-----------------------------
# A. point-level soil + disturbance table
#-----------------------------
soil_vars <- c("lbc","lbc_eq","p_h_2","ca","k","mg","mn","p","zn")
# Disturbance raster (UTM 17N)
current_dist <- rast("../data/disturbance/currently_dist_binary.tif")

# Recode categorical -> binary (you already did this)
lev <- levels(current_dist)[[1]]
id_col <- names(lev)[1]
yes_id <- lev[[id_col]][lev$CurrentDis == "Yes"]
current_dist_bin <- ifel(current_dist == yes_id, 1, 0)
names(current_dist_bin) <- "dist_current"

# OPTIONAL: distance-to-disturbance raster (meters)
# (distance from each cell to nearest disturbed cell)
dist_to_dist <- distance(current_dist_bin, value = 1)
names(dist_to_dist) <- "dist_to_current_m"

# --- disturbance rasters already made ---
# current_dist_bin (0/1) and dist_to_dist (meters)

# points in raster CRS
soil_pts_ll <- surface_sampling %>%
  select(row_num, x, y, lbc, lbc_eq, p_h_2, ca, k, mg, mn, p, zn) %>%
  st_as_sf(coords = c("x","y"), crs = 4326)

soil_pts_utm <- st_transform(soil_pts_ll, crs(current_dist_bin))

# extract disturbance values
dist_vals <- terra::extract(current_dist_bin, vect(soil_pts_utm)) %>% select(-ID)
dist_dvals <- terra::extract(dist_to_dist, vect(soil_pts_utm)) %>% select(-ID)

soil_dist <- soil_pts_utm %>%
  st_drop_geometry() %>%
  bind_cols(dist_vals, dist_dvals) %>%
  clean_names()




# current_dist_bin is 0/1
# make a target raster: 1 where disturbed, NA elsewhere
target <- ifel(current_dist_bin == 1, 1, NA)

# distance to nearest target cell (meters, since UTM)
dist_to_dist <- distance(target)
names(dist_to_dist) <- "dist_to_current_m"

plot(dist_to_dist)
global(dist_to_dist, "range", na.rm=TRUE)
dist_dvals <- terra::extract(dist_to_dist, vect(soil_pts_utm)) %>% select(-ID)

soil_dist <- soil_pts_utm %>%
  st_drop_geometry() %>%
  bind_cols(dist_vals, dist_dvals) %>%
  janitor::clean_names()

point_base <- soil_dist %>%
  mutate(row_num = as.character(row_num)) %>%
  clean_names()

#-----------------------------
# B. add diversity indices if you want them in the same table
#-----------------------------
idx_keep <- surface_sampling_idx %>%
  mutate(row_num = as.character(row_num)) %>%
  clean_names() %>%
  select(
    row_num,
    simpsons_diversity_index,
    shannon_weiner_diversity_index,
    pielous_evenness_index,
    berger_parker_dominance_index
  )

#-----------------------------
# C. choose any spatial covariates already living in your point table
#   edit this list to match your actual column names
#-----------------------------
candidate_covars <- c(
  "elevation", "dem", "slope", "aspect", "twi", "ndvi",
  "spi", "tri", "roughness", "curvature"
)

covar_vars <- intersect(candidate_covars, names(point_base))

#-----------------------------
# D. final point table (soil + disturbance + indices + spatial covariates)
#-----------------------------
point_env <- point_base %>%
  left_join(idx_keep, by = "row_num") %>%
  select(
    row_num,
    dist_current,
    dist_to_current_m,
    all_of(intersect(soil_vars, names(.))),
    all_of(covar_vars),
    simpsons_diversity_index,
    shannon_weiner_diversity_index,
    pielous_evenness_index,
    berger_parker_dominance_index
  )

glimpse(point_env)
```

```{r}
weed_trait_long <- surface_sampling_weed %>%
  clean_names() %>%
  transmute(
    row_num      = as.character(row_num),
    weed_species = str_squish(as.character(weed_species)),
    family       = str_squish(as.character(family)),
    genus        = str_squish(as.character(genus)),
    species      = str_squish(as.character(species)),
    life_cycle   = str_squish(as.character(life_cycle)),
    season       = str_squish(as.character(season)),
    type         = str_squish(as.character(type)),
    weed_count   = as.numeric(weed_count)
  ) %>%
  filter(!is.na(weed_count))

# standardize a bit
weed_trait_long <- weed_trait_long %>%
  mutate(
    life_cycle = case_when(
      str_detect(str_to_lower(life_cycle), "annual") ~ "Annual",
      str_detect(str_to_lower(life_cycle), "perennial") ~ "Perennial",
      TRUE ~ NA_character_
    ),
    type = case_when(
      str_detect(str_to_lower(type), "broad") ~ "Broadleaf",
      str_detect(str_to_lower(type), "grass") ~ "Grass",
      str_detect(str_to_lower(type), "sedge") ~ "Sedge",
      TRUE ~ NA_character_
    ),
    season = case_when(
      str_detect(str_to_lower(season), "warm") ~ "Warm",
      str_detect(str_to_lower(season), "cool") ~ "Cool",
      TRUE ~ NA_character_
    )
  )

# basic total/richness summaries
weed_point_summary <- weed_trait_long %>%
  group_by(row_num) %>%
  summarise(
    weed_total_abundance  = sum(weed_count, na.rm = TRUE),
    weed_species_richness = sum(weed_count > 0, na.rm = TRUE),
    weed_family_richness  = n_distinct(family[weed_count > 0]),
    .groups = "drop"
  )

# life cycle abundance
life_cycle_wide <- weed_trait_long %>%
  filter(!is.na(life_cycle)) %>%
  group_by(row_num, life_cycle) %>%
  summarise(abund = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = life_cycle,
    values_from = abund,
    values_fill = 0,
    names_prefix = "life_"
  )

# type abundance
type_wide <- weed_trait_long %>%
  filter(!is.na(type)) %>%
  group_by(row_num, type) %>%
  summarise(abund = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = type,
    values_from = abund,
    values_fill = 0,
    names_prefix = "type_"
  )

# season abundance
season_wide <- weed_trait_long %>%
  filter(!is.na(season)) %>%
  group_by(row_num, season) %>%
  summarise(abund = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = season,
    values_from = abund,
    values_fill = 0,
    names_prefix = "season_"
  )

# stacked trait groups: annual broadleaf, perennial grass, etc.
stacked_trait_wide <- weed_trait_long %>%
  mutate(
    stacked_trait = case_when(
      !is.na(life_cycle) & !is.na(type) ~ paste(life_cycle, type, sep = "_"),
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(stacked_trait)) %>%
  group_by(row_num, stacked_trait) %>%
  summarise(abund = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = stacked_trait,
    values_from = abund,
    values_fill = 0,
    names_prefix = "trait_"
  )

# combine all point-level weed summaries
weed_traits_point <- weed_point_summary %>%
  left_join(life_cycle_wide, by = "row_num") %>%
  left_join(type_wide, by = "row_num") %>%
  left_join(season_wide, by = "row_num") %>%
  left_join(stacked_trait_wide, by = "row_num") %>%
  clean_names()

glimpse(weed_traits_point)

analysis_dat <- point_env %>%
  left_join(weed_traits_point, by = "row_num") %>%
  mutate(
    dist_current = as.factor(dist_current)
  )

glimpse(analysis_dat)
```

```{r}
trait_vars <- names(analysis_dat) %>%
  keep(~ str_detect(.x, "^life_|^type_|^season_|^trait_|weed_total_abundance|weed_species_richness|weed_family_richness|simpsons|shannon|pielous|berger"))

df_long_traits <- analysis_dat %>%
  pivot_longer(all_of(trait_vars), names_to = "trait_var", values_to = "value") %>%
  mutate(value = as.numeric(value))

p_trait <- df_long_traits %>%
  filter(!is.na(value), !is.na(dist_current)) %>%
  group_by(trait_var) %>%
  summarise(
    n0 = sum(dist_current == "0"),
    n1 = sum(dist_current == "1"),
    p_value = ifelse(n0 >= 2 & n1 >= 2,
                     suppressWarnings(wilcox.test(value ~ dist_current)$p.value),
                     NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    p_bh = p.adjust(p_value, method = "BH"),
    p_lab = ifelse(is.na(p_value), "p=NA",
                   paste0("p=", signif(p_value, 3), " | BH=", signif(p_bh, 3)))
  )

plot_dat <- df_long_traits %>%
  left_join(
    df_long_traits %>%
      group_by(trait_var) %>%
      summarise(y_pos = max(value, na.rm = TRUE) * 1.05, .groups = "drop"),
    by = "trait_var"
  )

p_trait_plot <- p_trait %>%
  left_join(plot_dat %>% distinct(trait_var, y_pos), by = "trait_var")

ggplot(plot_dat, aes(x = dist_current, y = value)) +
  geom_boxplot(outlier.alpha = 0.25) +
  geom_jitter(width = 0.15, alpha = 0.25) +
  facet_wrap(~ trait_var, scales = "free_y") +
  geom_text(
    data = p_trait_plot,
    aes(x = 1.5, y = y_pos, label = p_lab),
    inherit.aes = FALSE,
    size = 2.8
  ) +
  labs(x = "Currently disturbed (0/1)", y = "Trait metric") +
  theme_minimal()
```
```{r}
ggplot(
  analysis_dat %>%
    pivot_longer(all_of(trait_vars), names_to = "trait_var", values_to = "value"),
  aes(x = dist_to_current_m, y = as.numeric(value))
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ trait_var, scales = "free_y") +
  labs(x = "Distance to current disturbance (m)", y = "Trait metric") +
  theme_minimal()
```
```{r}
dist_trait_cor <- analysis_dat %>%
  summarise(across(all_of(trait_vars), ~ suppressWarnings(
    cor(as.numeric(.x), dist_to_current_m, method = "spearman", use = "pairwise.complete.obs")
  ))) %>%
  pivot_longer(everything(), names_to = "trait_var", values_to = "rho") %>%
  arrange(desc(abs(rho)))

dist_trait_cor
```
```{r}
predictor_vars <- c(
  intersect(c("dist_current", "dist_to_current_m"), names(analysis_dat)),
  intersect(soil_vars, names(analysis_dat)),
  covar_vars
)

predictor_vars_num <- predictor_vars[predictor_vars != "dist_current"]

cor_grid <- expand_grid(
  trait_var = trait_vars,
  predictor = predictor_vars_num
) %>%
  rowwise() %>%
  mutate(
    rho = suppressWarnings(
      cor(
        as.numeric(analysis_dat[[trait_var]]),
        as.numeric(analysis_dat[[predictor]]),
        method = "spearman",
        use = "pairwise.complete.obs"
      )
    )
  ) %>%
  ungroup()

ggplot(cor_grid, aes(x = predictor, y = trait_var, fill = rho)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(rho, 2)), size = 2.6) +
  scale_fill_viridis_c(option = "magma", limits = c(-1, 1), name = "Spearman\nrho") +
  labs(x = "Environmental predictor", y = "Weed trait metric") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```
```{r}
rf_predictors <- c(
  intersect(soil_vars, names(analysis_dat)),
  intersect(covar_vars, names(analysis_dat)),
  "dist_to_current_m"
)

rf_predictors <- unique(rf_predictors)

rf_targets <- c(
  "weed_total_abundance",
  "weed_species_richness",
  "weed_family_richness",
  "shannon_weiner_diversity_index",
  "simpsons_diversity_index",
  "life_annual",
  "life_perennial",
  "type_broadleaf",
  "type_grass",
  "type_sedge"
)

rf_targets <- intersect(rf_targets, names(analysis_dat))

fit_rf_importance <- function(target_var, dat, predictors) {
  d <- dat %>%
    select(all_of(c(target_var, predictors))) %>%
    mutate(across(everything(), as.numeric)) %>%
    drop_na()

  if (nrow(d) < 15) return(NULL)

  mod <- ranger(
    formula = as.formula(paste(target_var, "~ .")),
    data = d,
    num.trees = 500,
    importance = "permutation",
    seed = 1
  )

  tibble(
    response = target_var,
    predictor = names(mod$variable.importance),
    importance = as.numeric(mod$variable.importance)
  )
}

rf_imp <- map_dfr(rf_targets, fit_rf_importance, dat = analysis_dat, predictors = rf_predictors)

ggplot(rf_imp, aes(x = predictor, y = response, fill = importance)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(importance, 2)), size = 2.6) +
  scale_fill_viridis_c() +
  labs(x = "Predictor", y = "Response") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```
```{r}
analysis_bins <- analysis_dat %>%
  mutate(
    dist_bin = cut(
      dist_to_current_m,
      breaks = c(-Inf, 5, 15, 30, Inf),
      labels = c("0-5 m", "5-15 m", "15-30 m", ">30 m")
    )
  )

dom_family_by_bin <- weed_trait_long %>%
  group_by(row_num, family) %>%
  summarise(fam_abund = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  group_by(row_num) %>%
  slice_max(fam_abund, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(analysis_bins %>% select(row_num, dist_bin), by = "row_num") %>%
  count(dist_bin, family) %>%
  group_by(dist_bin) %>%
  mutate(pct = n / sum(n))

ggplot(dom_family_by_bin, aes(x = dist_bin, y = pct, fill = family)) +
  geom_col() +
  scale_y_continuous(labels = percent) +
  labs(x = "Distance to disturbance", y = "Proportion of points", fill = "Dominant family") +
  theme_minimal()


# weed long table
weed_family_long <- surface_sampling_weed %>%
  janitor::clean_names() %>%
  transmute(
    row_num    = as.character(row_num),
    family     = str_squish(as.character(family)),
    weed_count = as.numeric(weed_count)
  ) %>%
  filter(!is.na(weed_count), weed_count > 0, !is.na(family), family != "")

# total abundance per point
point_totals <- weed_family_long %>%
  group_by(row_num) %>%
  summarise(total_abundance = sum(weed_count, na.rm = TRUE), .groups = "drop")

# family abundance per point
family_point <- weed_family_long %>%
  group_by(row_num, family) %>%
  summarise(family_abundance = sum(weed_count, na.rm = TRUE), .groups = "drop") %>%
  left_join(point_totals, by = "row_num") %>%
  mutate(
    family_prop = family_abundance / total_abundance
  )

family_point_dist <- family_point_dist %>%
  filter(total_abundance >= 3)

top_fams <- family_point_dist %>%
  group_by(family) %>%
  summarise(total_family_abundance = sum(family_abundance, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_family_abundance, n = 6) %>%
  pull(family)

family_model_dat <- family_point_dist %>%
  filter(family %in% top_fams) %>%
  mutate(other_abundance = total_abundance - family_abundance)
family_plot_dat <- family_point_dist %>%
  filter(family %in% top_fams)

ggplot(family_plot_dat, aes(x = dist_to_current_m, y = family_prop, color = family)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Distance to disturbance (m)",
    y = "Family proportion of total weeds",
    color = "Family"
  ) +
  theme_minimal()

family_bin_dat <- family_plot_dat %>%
  mutate(
    dist_bin = cut_width(dist_to_current_m, width = 5, boundary = 0)
  ) %>%
  group_by(family, dist_bin) %>%
  summarise(
    mid_dist = mean(dist_to_current_m, na.rm = TRUE),
    mean_prop = mean(family_prop, na.rm = TRUE),
    se_prop = sd(family_prop, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 3)

ggplot(family_bin_dat, aes(x = mid_dist, y = mean_prop, color = family)) +
  geom_line() +
  geom_point() +
  geom_ribbon(
    aes(ymin = pmax(mean_prop - se_prop, 0),
        ymax = pmin(mean_prop + se_prop, 1),
        fill = family),
    alpha = 0.10,
    color = NA
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Distance to disturbance (m)",
    y = "Mean family proportion",
    color = "Family",
    fill = "Family"
  ) +
  theme_minimal()

```

```{r}
library(tidyverse)
library(mgcv)
library(broom)

family_gam_results <- family_point_dist %>%
  filter(family %in% top_fams) %>%
  mutate(other_abundance = total_abundance - family_abundance) %>%
  group_by(family) %>%
  group_modify(~{
    d <- .x %>%
      filter(
        total_abundance > 0,
        is.finite(dist_to_current_m),
        is.finite(family_abundance),
        is.finite(other_abundance)
      )

    if(nrow(d) < 20) return(tibble())

    mod <- gam(
      cbind(family_abundance, other_abundance) ~ s(dist_to_current_m, k = 4),
      data = d,
      family = quasibinomial(),
      method = "REML"
    )

    sm <- summary(mod)

    tibble(
      family = unique(d$family),
      n = nrow(d),
      edf = sm$s.table[1, "edf"],
      F = sm$s.table[1, "F"],
      p_value = sm$s.table[1, "p-value"]
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_bh = p.adjust(p_value, method = "BH"),
    nonlinear = edf > 1
  ) %>%
  arrange(p_bh)

family_gam_results
```
###### Spatial linkages for disturbance and family

```{r}

library(mgcv)
family_point_dist <- family_point %>%
  left_join(
    point_env %>% select(row_num, dist_to_current_m),
    by = "row_num"
  ) %>%
  left_join(
    coord_lookup,
    by = "row_num"
  ) %>%
  filter(
    !is.na(dist_to_current_m),
    !is.na(x_utm),
    !is.na(y_utm)
  )
family_model_dat <- family_point_dist %>%
  filter(total_abundance >= 3) %>%
  group_by(family) %>%
  mutate(total_family_abundance = sum(family_abundance, na.rm = TRUE)) %>%
  ungroup()

top_fams <- family_model_dat %>%
  distinct(family, total_family_abundance) %>%
  slice_max(total_family_abundance, n = 6) %>%
  pull(family)

family_model_dat <- family_model_dat %>%
  filter(family %in% top_fams) %>%
  mutate(other_abundance = total_abundance - family_abundance)

glimpse(family_model_dat)
glimpse(family_point_dist)

library(mgcv)
library(tibble)

fit_family_spatial_gams <- function(d, key) {
  fam_name <- key$family[1]

  d <- d %>%
    filter(
      total_abundance > 0,
      is.finite(dist_to_current_m),
      is.finite(x_utm),
      is.finite(y_utm),
      is.finite(family_abundance),
      is.finite(other_abundance)
    )

  if (nrow(d) < 25) {
    return(tibble(
      family = fam_name,
      n = nrow(d),
      aic_m0 = NA_real_,
      aic_m1 = NA_real_,
      aic_m2 = NA_real_,
      aic_m3 = NA_real_,
      dev_exp_m1 = NA_real_,
      dev_exp_m2 = NA_real_,
      dev_exp_m3 = NA_real_,
      edf_dist_m1 = NA_real_,
      p_dist_m1 = NA_real_,
      edf_dist_m3 = NA_real_,
      p_dist_m3 = NA_real_,
      edf_space_m3 = NA_real_,
      p_space_m3 = NA_real_
    ))
  }

  m0 <- gam(
    cbind(family_abundance, other_abundance) ~ 1,
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m1 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(dist_to_current_m, k = 4),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m2 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(x_utm, y_utm, k = 15),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m3 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(dist_to_current_m, k = 4) +
      s(x_utm, y_utm, k = 15),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  sm1 <- summary(m1)
  sm2 <- summary(m2)
  sm3 <- summary(m3)

  tibble(
    family = fam_name,
    n = nrow(d),

    aic_m0 = AIC(m0),
    aic_m1 = AIC(m1),
    aic_m2 = AIC(m2),
    aic_m3 = AIC(m3),

    dev_exp_m1 = sm1$dev.expl,
    dev_exp_m2 = sm2$dev.expl,
    dev_exp_m3 = sm3$dev.expl,

    edf_dist_m1 = sm1$s.table[1, "edf"],
    p_dist_m1   = sm1$s.table[1, "p-value"],

    edf_dist_m3 = sm3$s.table[1, "edf"],
    p_dist_m3   = sm3$s.table[1, "p-value"],

    edf_space_m3 = sm3$s.table[2, "edf"],
    p_space_m3   = sm3$s.table[2, "p-value"]
  )
}
fit_family_spatial_gams <- function(d) {
  fam_name <- unique(d$family)

  d <- d %>%
    filter(
      total_abundance > 0,
      is.finite(dist_to_current_m),
      is.finite(x_utm),
      is.finite(y_utm),
      is.finite(family_abundance),
      is.finite(other_abundance)
    )

  if (nrow(d) < 25) return(NULL)

  m0 <- gam(
    cbind(family_abundance, other_abundance) ~ 1,
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m1 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(dist_to_current_m, k = 4),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m2 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(x_utm, y_utm, k = 20),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m3 <- gam(
    cbind(family_abundance, other_abundance) ~
      s(dist_to_current_m, k = 4) +
      s(x_utm, y_utm, k = 20),
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  sm3 <- summary(m3)

  tibble(
    family = fam_name,
    n = nrow(d),

    aic_m0 = AIC(m0),
    aic_m1 = AIC(m1),
    aic_m2 = AIC(m2),
    aic_m3 = AIC(m3),

    dev_exp_m1 = summary(m1)$dev.expl,
    dev_exp_m2 = summary(m2)$dev.expl,
    dev_exp_m3 = summary(m3)$dev.expl,

    edf_dist_m1 = summary(m1)$s.table[1, "edf"],
    p_dist_m1   = summary(m1)$s.table[1, "p-value"],

    edf_dist_m3 = sm3$s.table[1, "edf"],
    p_dist_m3   = sm3$s.table[1, "p-value"],

    edf_space_m3 = sm3$s.table[2, "edf"],
    p_space_m3   = sm3$s.table[2, "p-value"]
  )
}

family_spatial_results <- family_model_dat %>%
  group_by(family) %>%
  group_modify(~ fit_family_spatial_gams(.x)) %>%
  ungroup() %>%
  mutate(
    p_dist_m3_bh = p.adjust(p_dist_m3, method = "BH"),
    p_space_m3_bh = p.adjust(p_space_m3, method = "BH"),
    nonlinear_dist = edf_dist_m3 > 1
  ) %>%
  arrange(p_dist_m3_bh)

family_spatial_results

```

```{r}
library(tidyverse)
library(mgcv)
library(janitor)
 
covariates <- rast("../data/covars/covariate_stack_3m.tif")
names(covariates)
names(covariates) <- make.names(names(covariates))
names(covariates)

soil_pts_cov <- terra::extract(covariates, vect(soil_pts_utm), bind = TRUE) %>%
  st_as_sf()

soil_pts_cov_df <- soil_pts_cov %>%
  st_drop_geometry() %>%
  clean_names() %>%
  mutate(row_num = as.character(row_num))

glimpse(soil_pts_cov_df)

soil_vars <- c("lbc", "lbc_eq", "p_h_2", "ca", "k", "mg", "mn", "p", "zn")

env_dat <- point_env %>%
  select(row_num, x_utm, y_utm, dist_current, dist_to_current_m, any_of(soil_vars)) %>%
  left_join(
    soil_pts_cov_df,
    by = "row_num",
    suffix = c("", "_cov")
  )

glimpse(env_dat)
candidate_predictors <- c(
  "dist_to_current_m",
  "dist_current",
  "p_h_2", "p", "k", "ca", "mg", "mn", "zn",
  "elevation",
  "slope",
  "topographic_wetness_index_usgs",
  "topographic_position_index",
  "flow_accumulation_usgs",
  "stream_power_index",
  "vertical_distance_to_channel_network",
  "ndvi_sept_2022"
)

candidate_predictors <- intersect(candidate_predictors, names(env_dat))
candidate_predictors

family_env_dat <- family_point %>%
  left_join(env_dat, by = "row_num") %>%
  mutate(
    other_abundance = total_abundance - family_abundance
  ) %>%
  filter(
    total_abundance >= 3,
    !is.na(x_utm),
    !is.na(y_utm)
  )

glimpse(family_env_dat)

num_preds <- family_env_dat %>%
  select(any_of(candidate_predictors)) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.x))))

cor_mat <- cor(num_preds, use = "pairwise.complete.obs", method = "spearman")
round(cor_mat, 2)

cbind(family_abundance, other_abundance) ~
  s(dist_to_current_m, k = 4) +
  p_h_2 + p + k + ca + mg +
  elevation + slope + topographic_wetness_index_usgs +
  flow_accumulation_usgs + ndvi_sept_2022 +
  s(x_utm, y_utm, k = 15)


library(mgcv)
library(broom)
library(tibble)

model_predictors <- c(
  "p_h_2", "p", "k", "ca", "mg",
  "elevation", "slope",
  "topographic_wetness_index_usgs",
  "flow_accumulation_usgs",
  "ndvi_sept_2022"
)

model_predictors <- intersect(model_predictors, names(family_env_dat))

fit_family_env_gam <- function(d, key, predictors) {
  fam_name <- key$family[1]

  keep_vars <- c(
    "family_abundance", "other_abundance", "dist_to_current_m",
    "x_utm", "y_utm", predictors
  )

  d <- d %>%
    dplyr::select(any_of(c("family", keep_vars))) %>%
    mutate(across(any_of(keep_vars), as.numeric)) %>%
    drop_na()

  if (nrow(d) < 30) {
    return(tibble(
      family = fam_name,
      n = nrow(d),
      dev_exp_env = NA_real_,
      dev_exp_env_space = NA_real_,
      edf_dist = NA_real_,
      p_dist = NA_real_,
      edf_space = NA_real_,
      p_space = NA_real_
    ))
  }

  rhs_env <- paste(
    c("s(dist_to_current_m, k = 4)", predictors),
    collapse = " + "
  )

  rhs_env_space <- paste(
    c("s(dist_to_current_m, k = 4)", predictors, "s(x_utm, y_utm, k = 15)"),
    collapse = " + "
  )

  f_env <- as.formula(
    paste("cbind(family_abundance, other_abundance) ~", rhs_env)
  )

  f_env_space <- as.formula(
    paste("cbind(family_abundance, other_abundance) ~", rhs_env_space)
  )

  m_env <- gam(
    f_env,
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  m_env_space <- gam(
    f_env_space,
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  sm_env <- summary(m_env)
  sm_env_space <- summary(m_env_space)

  # parametric terms from full model
  param_tab <- broom::tidy(m_env_space) %>%
    filter(term != "(Intercept)") %>%
    mutate(family = fam_name)

  smooth_tab <- tibble(
    family = fam_name,
    n = nrow(d),
    dev_exp_env = sm_env$dev.expl,
    dev_exp_env_space = sm_env_space$dev.expl,
    edf_dist = sm_env_space$s.table[1, "edf"],
    p_dist = sm_env_space$s.table[1, "p-value"],
    edf_space = sm_env_space$s.table[nrow(sm_env_space$s.table), "edf"],
    p_space = sm_env_space$s.table[nrow(sm_env_space$s.table), "p-value"]
  )

  list(
    smooth = smooth_tab,
    param = param_tab
  )
}

top_fams <- family_env_dat %>%
  group_by(family) %>%
  summarise(total_family_abundance = sum(family_abundance, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_family_abundance, n = 6) %>%
  pull(family)

family_env_top <- family_env_dat %>%
  filter(family %in% top_fams)

fam_split <- family_env_top %>%
  group_split(family)

fam_keys <- family_env_top %>%
  group_keys(family)

env_results_list <- map2(
  fam_split,
  split(fam_keys, seq_len(nrow(fam_keys))),
  ~ fit_family_env_gam(.x, .y, predictors = model_predictors)
)

family_env_smooth <- map_dfr(env_results_list, "smooth") %>%
  mutate(
    p_dist_bh = p.adjust(p_dist, method = "BH"),
    p_space_bh = p.adjust(p_space, method = "BH"),
    nonlinear_dist = edf_dist > 1
  )

family_env_param <- map_dfr(env_results_list, "param") %>%
  group_by(term) %>%
  mutate(p_adj_bh = p.adjust(p.value, method = "BH")) %>%
  ungroup()

family_env_smooth
family_env_param
```

```{r}
library(tidyverse)
library(mgcv)
library(tibble)

model_predictors <- c(
  "p_h_2", "p", "k", "ca",
  "elevation", "slope",
  "topographic_wetness_index_usgs",
  "topographic_position_index",
  "ndvi_sept_2022"
)

fit_family_env_gam <- function(d, fam_name, predictors) {

  keep_vars <- c(
    "family_abundance", "other_abundance",
    "dist_to_current_m", "x_utm", "y_utm",
    predictors
  )

  d2 <- d %>%
    select(any_of(keep_vars)) %>%
    mutate(across(everything(), as.numeric)) %>%
    drop_na()

  if (nrow(d2) < 30) {
    return(list(
      smooth = tibble(
        family = fam_name,
        n = nrow(d2),
        dev_exp_env = NA_real_,
        dev_exp_env_space = NA_real_,
        edf_dist = NA_real_,
        p_dist = NA_real_,
        edf_space = NA_real_,
        p_space = NA_real_
      ),
      param = tibble(
        family = character(),
        term = character(),
        estimate = numeric(),
        std_error = numeric(),
        statistic = numeric(),
        p_value = numeric()
      )
    ))
  }

  f_env <- as.formula(
    paste(
      "cbind(family_abundance, other_abundance) ~ s(dist_to_current_m, k = 4) +",
      paste(predictors, collapse = " + ")
    )
  )

  f_env_space <- as.formula(
    paste(
      "cbind(family_abundance, other_abundance) ~ s(dist_to_current_m, k = 4) +",
      paste(predictors, collapse = " + "),
      "+ s(x_utm, y_utm, k = 15)"
    )
  )

  m_env <- gam(
    f_env,
    data = d2,
    family = quasibinomial(),
    method = "REML"
  )

  m_env_space <- gam(
    f_env_space,
    data = d2,
    family = quasibinomial(),
    method = "REML"
  )

  sm_env <- summary(m_env)
  sm_env_space <- summary(m_env_space)

  smooth_tab <- tibble(
    family = fam_name,
    n = nrow(d2),
    dev_exp_env = sm_env$dev.expl,
    dev_exp_env_space = sm_env_space$dev.expl,
    edf_dist = sm_env_space$s.table[1, "edf"],
    p_dist = sm_env_space$s.table[1, "p-value"],
    edf_space = sm_env_space$s.table[2, "edf"],
    p_space = sm_env_space$s.table[2, "p-value"]
  )

  # safer extraction of parametric terms
  ptab <- as.data.frame(sm_env_space$p.table)

  if (nrow(ptab) == 0) {
    param_tab <- tibble(
      family = character(),
      term = character(),
      estimate = numeric(),
      std_error = numeric(),
      statistic = numeric(),
      p_value = numeric()
    )
  } else {
    ptab$term <- rownames(ptab)
    rownames(ptab) <- NULL

    param_tab <- ptab %>%
      as_tibble() %>%
      rename(
        estimate = Estimate,
        std_error = `Std. Error`,
        statistic = `t value`,
        p_value = `Pr(>|t|)`
      ) %>%
      filter(term != "(Intercept)") %>%
      mutate(family = fam_name) %>%
      select(family, term, estimate, std_error, statistic, p_value)
  }

  list(
    smooth = smooth_tab,
    param = param_tab
  )
}

top_fams <- family_env_dat %>%
  group_by(family) %>%
  summarise(total_family_abundance = sum(family_abundance, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_family_abundance, n = 6) %>%
  pull(family)

env_results_list <- top_fams %>%
  set_names() %>%
  map(~{
    d <- family_env_dat %>%
      filter(family == .x)
    fit_family_env_gam(d, fam_name = .x, predictors = model_predictors)
  })

family_env_smooth <- map_dfr(env_results_list, "smooth") %>%
  mutate(
    p_dist_bh = p.adjust(p_dist, method = "BH"),
    p_space_bh = p.adjust(p_space, method = "BH"),
    nonlinear_dist = edf_dist > 1
  ) %>%
  arrange(p_space_bh, p_dist_bh)

family_env_param <- map_dfr(env_results_list, "param") %>%
  group_by(term) %>%
  mutate(p_adj_bh = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  arrange(term, p_adj_bh)

family_env_smooth
family_env_param
sm_test <- summary(
  gam(
    cbind(family_abundance, other_abundance) ~
      s(dist_to_current_m, k = 4) + p_h_2 + p + k + ca +
      elevation + slope + topographic_wetness_index_usgs +
      topographic_position_index + ndvi_sept_2022 +
      s(x_utm, y_utm, k = 15),
    data = family_env_dat %>% filter(family == top_fams[1]) %>% drop_na(),
    family = quasibinomial(),
    method = "REML"
  )
)

colnames(as.data.frame(sm_test$p.table))

family_env_param %>%
  arrange(p_adj_bh, p_value)
family_env_param %>%
  filter(p_adj_bh < 0.05) %>%
  arrange(term, p_adj_bh)

```

```{r}
#cyperaceae and poaceae

fit_full_family_model <- function(fam_name, dat, predictors) {
  d <- dat %>%
    filter(family == fam_name) %>%
    mutate(other_abundance = total_abundance - family_abundance) %>%
    select(any_of(c(
      "family_abundance", "other_abundance",
      "dist_to_current_m", "x_utm", "y_utm",
      predictors
    ))) %>%
    mutate(across(everything(), as.numeric)) %>%
    drop_na()

  f_env_space <- as.formula(
    paste(
      "cbind(family_abundance, other_abundance) ~ s(dist_to_current_m, k = 4) +",
      paste(predictors, collapse = " + "),
      "+ s(x_utm, y_utm, k = 15)"
    )
  )

  mod <- gam(
    f_env_space,
    data = d,
    family = quasibinomial(),
    method = "REML"
  )

  list(data = d, model = mod)
}

make_partial_distance_pred <- function(fam_name, dat, predictors) {
  out <- fit_full_family_model(fam_name, dat, predictors)

  d <- out$data
  mod <- out$model

  newdat <- tibble(
    dist_to_current_m = seq(
      min(d$dist_to_current_m, na.rm = TRUE),
      max(d$dist_to_current_m, na.rm = TRUE),
      length.out = 200
    ),
    x_utm = median(d$x_utm, na.rm = TRUE),
    y_utm = median(d$y_utm, na.rm = TRUE)
  )

  for (v in predictors) {
    newdat[[v]] <- median(d[[v]], na.rm = TRUE)
  }

  pr <- predict(mod, newdata = newdat, type = "link", se.fit = TRUE)

  newdat %>%
    mutate(
      family = fam_name,
      fit = plogis(pr$fit),
      lower = plogis(pr$fit - 1.96 * pr$se.fit),
      upper = plogis(pr$fit + 1.96 * pr$se.fit)
    )
}

partial_preds <- bind_rows(
  make_partial_distance_pred("Poaceae", family_env_dat, model_predictors),
  make_partial_distance_pred("Cyperaceae", family_env_dat, model_predictors)
)

ggplot(partial_preds, aes(x = dist_to_current_m, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  facet_wrap(~ family, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Distance to disturbance (m)",
    y = "Predicted family proportion\n(holding other covariates constant)"
  ) +
  theme_minimal()

plot_points <- family_env_dat %>%
  filter(family %in% c("Poaceae", "Cyperaceae")) %>%
  mutate(family_prop = family_abundance / total_abundance)

ggplot() +
  geom_point(
    data = plot_points,
    aes(x = dist_to_current_m, y = family_prop),
    alpha = 0.18,
    size = 1
  ) +
  geom_ribbon(
    data = partial_preds,
    aes(x = dist_to_current_m, ymin = lower, ymax = upper),
    alpha = 0.2
  ) +
  geom_line(
    data = partial_preds,
    aes(x = dist_to_current_m, y = fit),
    linewidth = 1
  ) +
  facet_wrap(~ family, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Distance to disturbance (m)",
    y = "Family proportion"
  ) +
  theme_minimal()

# plot_smooth_term <- function(fam_name, dat, predictors) {
#   out <- fit_full_family_model(fam_name, dat, predictors)
#   d <- out$data
#   mod <- out$model
# 
#   newdat <- tibble(
#     dist_to_current_m = seq(
#       min(d$dist_to_current_m, na.rm = TRUE),
#       max(d$dist_to_current_m, na.rm = TRUE),
#       length.out = 200
#     ),
#     x_utm = median(d$x_utm, na.rm = TRUE),
#     y_utm = median(d$y_utm, na.rm = TRUE)
#   )
# 
#   for (v in predictors) {
#     newdat[[v]] <- median(d[[v]], na.rm = TRUE)
#   }
# 
#   term_pred <- predict(
#     mod,
#     newdata = newdat,
#     type = "terms",
#     se.fit = TRUE
#   )
# 
#   term_name <- grep("dist_to_current_m", colnames(term_pred$fit), value = TRUE)
# 
#   tibble(
#     family = fam_name,
#     dist_to_current_m = newdat$dist_to_current_m,
#     fit = term_pred$fit[, term_name],
#     se = term_pred$se.fit[, term_name],
#     lower = fit - 1.96 * se,
#     upper = fit + 1.96 * se
#   )
# }
# 
# smooth_terms <- bind_rows(
#   plot_smooth_term("Poaceae", family_env_dat, model_predictors),
#   plot_smooth_term("Cyperaceae", family_env_dat, model_predictors)
# )
# 
# ggplot(smooth_terms, aes(x = dist_to_current_m, y = fit)) +
#   geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
#   geom_line(linewidth = 1) +
#   facet_wrap(~ family, scales = "free_y") +
#   labs(
#     x = "Distance to disturbance (m)",
#     y = "Partial smooth effect on link scale"
#   ) +
#   theme_minimal()

ggplot(
  plot_points,
  aes(x = dist_to_current_m, y = family_prop)
) +
  geom_point(alpha = 0.18, size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4), se = TRUE) +
  facet_wrap(~ family, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal()

#poaceae:
ggplot(
  family_env_dat %>% filter(family == "Poaceae"),
  aes(x = dist_to_current_m)
) +
  geom_histogram(bins = 20) +
  theme_minimal()

#sparse 200-400, poor structure and fit. cannot signal disturbance

#cyperaceae:

cyper_dat <- family_env_dat %>%
  filter(family == "Cyperaceae") %>%
  mutate(other_abundance = total_abundance - family_abundance) %>%
  select(
    family_abundance, other_abundance,
    dist_to_current_m, x_utm, y_utm,
    p_h_2, p, k, ca,
    elevation, slope,
    topographic_wetness_index_usgs,
    topographic_position_index,
    ndvi_sept_2022
  ) %>%
  mutate(across(everything(), as.numeric)) %>%
  drop_na()
m1 <- gam(
  cbind(family_abundance, other_abundance) ~
    s(dist_to_current_m, k = 4) +
    p_h_2 + p + k + ca +
    elevation + slope +
    topographic_wetness_index_usgs +
    topographic_position_index +
    ndvi_sept_2022 +
    s(x_utm, y_utm, k = 15),
  data = cyper_dat,
  family = quasibinomial(),
  method = "REML"
)

summary(m1)

m2 <- gam(
  cbind(family_abundance, other_abundance) ~
    s(dist_to_current_m, k = 4) +
    p_h_2 + p + k + ca +
    elevation + slope +
    s(topographic_wetness_index_usgs, k = 4) +
    topographic_position_index +
    ndvi_sept_2022 +
    s(x_utm, y_utm, k = 15),
  data = cyper_dat,
  family = quasibinomial(),
  method = "REML"
)

summary(m2)

m3 <- gam(
  cbind(family_abundance, other_abundance) ~
    s(dist_to_current_m, k = 4) +
    p_h_2 + p + k + ca +
    elevation + slope +
    s(topographic_wetness_index_usgs, k = 4) +
    s(topographic_position_index, k = 4) +
    ndvi_sept_2022 +
    s(x_utm, y_utm, k = 15),
  data = cyper_dat,
  family = quasibinomial(),
  method = "REML"
)

summary(m3)
m4 <- gam(
  cbind(family_abundance, other_abundance) ~
    p_h_2 + p + k + ca +
    elevation + slope +
    s(topographic_wetness_index_usgs, k = 4) +
    s(topographic_position_index, k = 4) +
    ndvi_sept_2022 +
    s(x_utm, y_utm, k = 15),
  data = cyper_dat,
  family = quasibinomial(),
  method = "REML"
)

summary(m4)

extract_gam_info <- function(mod, name) {
  sm <- summary(mod)

  tibble(
    model = name,
    dev_exp = sm$dev.expl,
    smooth = rownames(sm$s.table),
    edf = sm$s.table[, "edf"],
    p_value = sm$s.table[, "p-value"]
  )
}

bind_rows(
  extract_gam_info(m1, "m1_baseline"),
  extract_gam_info(m2, "m2_wetness_smooth"),
  extract_gam_info(m3, "m3_wetness_tpi_smooth"),
  extract_gam_info(m4, "m4_no_distance")
)
#distance:
plot(m3, select = 1, shade = TRUE, pages = 1)
#wetness:
plot(m3, select = 2, shade = TRUE, pages = 1)

make_partial_pred <- function(mod, dat, focal_var, grid_n = 200) {
  newdat <- dat[rep(1, grid_n), ]

  for (nm in names(newdat)) {
    if (is.numeric(newdat[[nm]])) {
      newdat[[nm]] <- median(dat[[nm]], na.rm = TRUE)
    }
  }

  newdat[[focal_var]] <- seq(
    min(dat[[focal_var]], na.rm = TRUE),
    max(dat[[focal_var]], na.rm = TRUE),
    length.out = grid_n
  )

  pr <- predict(mod, newdata = newdat, type = "link", se.fit = TRUE)

  tibble(
    x = newdat[[focal_var]],
    fit = plogis(pr$fit),
    lower = plogis(pr$fit - 1.96 * pr$se.fit),
    upper = plogis(pr$fit + 1.96 * pr$se.fit),
    variable = focal_var
  )
}

cyper_dist_pred <- make_partial_pred(m3, cyper_dat, "dist_to_current_m")
cyper_twi_pred  <- make_partial_pred(m3, cyper_dat, "topographic_wetness_index_usgs")

ggplot(cyper_dist_pred, aes(x = x, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  labs(
    x = "Distance to disturbance (m)",
    y = "Predicted Cyperaceae proportion"
  ) +
  theme_minimal()

ggplot(cyper_twi_pred, aes(x = x, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  labs(
    x = "Topographic Wetness Index",
    y = "Predicted Cyperaceae proportion"
  ) +
  theme_minimal()

#plot map of cyperaceae abundance across field site:

plot_dat <- family_env_dat %>%
  filter(family == "Cyperaceae") %>%
  mutate(family_prop = family_abundance / total_abundance)
ggplot(plot_dat, aes(x = x_utm, y = y_utm, color = family_prop)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_c() +  
  labs(
    x = "Easting (m)",
    y = "Northing (m)",
    color = "Cyperaceae proportion"
  ) +
  theme_minimal()
#Cyperaceae were concentrated nearer disturbance features, but this pattern was not clearly separable from hydrologic structure because many disturbances occurred in stream-adjacent areas. When disturbance distance was omitted, topographic wetness became a significant nonlinear predictor, indicating that Cyperaceae were likely responding to a coupled disturbance-moisture gradient rather than disturbance alone.

cyper_dat %>%
  summarise(
    rho = cor(dist_to_current_m, topographic_wetness_index_usgs,
              method = "spearman", use = "pairwise.complete.obs")
  )
# distance and wetness are not correlated, rho -0.02, so the fact that both are significant in the model suggests they are capturing different aspects of the environment that influence Cyperaceae abundance.

#-----------------------------
# helper: partial prediction
#-----------------------------
make_partial_pred <- function(mod, dat, focal_var, grid_n = 200) {
  newdat <- dat[rep(1, grid_n), , drop = FALSE]

  # hold numeric predictors constant at medians
  for (nm in names(newdat)) {
    if (is.numeric(newdat[[nm]])) {
      newdat[[nm]] <- median(dat[[nm]], na.rm = TRUE)
    }
  }

  # vary focal predictor across observed range
  newdat[[focal_var]] <- seq(
    min(dat[[focal_var]], na.rm = TRUE),
    max(dat[[focal_var]], na.rm = TRUE),
    length.out = grid_n
  )

  pr <- predict(mod, newdata = newdat, type = "link", se.fit = TRUE)

  tibble(
    x = newdat[[focal_var]],
    fit = plogis(pr$fit),
    lower = plogis(pr$fit - 1.96 * pr$se.fit),
    upper = plogis(pr$fit + 1.96 * pr$se.fit),
    variable = focal_var
  )
}

#-----------------------------
# build plotting data
#-----------------------------
cyper_plot_raw <- family_env_dat %>%
  filter(family == "Cyperaceae") %>%
  mutate(
    family_prop = family_abundance / total_abundance
  )

cyper_dist_pred <- make_partial_pred(m3, cyper_dat, "dist_to_current_m")
cyper_twi_pred  <- make_partial_pred(m3, cyper_dat, "topographic_wetness_index_usgs")

#-----------------------------
# common theme
#-----------------------------
theme_cyper <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 10, hjust = 0),
    legend.position = "none"
  )

#-----------------------------
# panel 1: distance effect
#-----------------------------
p_dist <- ggplot() +
  geom_point(
    data = cyper_plot_raw,
    aes(x = dist_to_current_m, y = family_prop),
    alpha = 0.18,
    size = 1.8
  ) +
  geom_ribbon(
    data = cyper_dist_pred,
    aes(x = x, ymin = lower, ymax = upper),
    alpha = 0.22
  ) +
  geom_line(
    data = cyper_dist_pred,
    aes(x = x, y = fit),
    linewidth = 1.1
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Distance to disturbance",
    x = "Distance to disturbance (m)",
    y = "Predicted Cyperaceae proportion"
  ) +
  theme_cyper

#-----------------------------
# panel 2: wetness effect
#-----------------------------
p_twi <- ggplot() +
  geom_point(
    data = cyper_plot_raw,
    aes(x = topographic_wetness_index_usgs, y = family_prop),
    alpha = 0.18,
    size = 1.8
  ) +
  geom_ribbon(
    data = cyper_twi_pred,
    aes(x = x, ymin = lower, ymax = upper),
    alpha = 0.22
  ) +
  geom_line(
    data = cyper_twi_pred,
    aes(x = x, y = fit),
    linewidth = 1.1
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Topographic wetness",
    x = "Topographic Wetness Index",
    y = NULL
  ) +
  theme_cyper


#-----------------------------
# combine panels
#-----------------------------
library(patchwork)
p_dist <- p_dist +
  annotate("text", x = Inf, y = Inf,
           label = "edf = 2.30\np = 0.0078",
           hjust = 1.1, vjust = 1.4, size = 4)

p_twi <- p_twi +
  annotate("text", x = Inf, y = Inf,
           label = "edf = 1.18\np = 0.639",
           hjust = 1.1, vjust = 1.4, size = 4)
cyper_fig <- p_dist + p_twi +
  plot_annotation(
    title = "Adjusted partial effects for Cyperaceae",
    subtitle = paste(
      "Lines show model-predicted Cyperaceae proportion from the spatial GAM;",
      "other covariates held at typical values."
    ),
    caption = paste(
      "Points are raw observed proportions for context only.",
      "Shaded ribbons show 95% confidence intervals."
    )
  )

cyper_fig


```










